---
title: "Regression models on Motor Trend cars dataset"
author: "Molina Rafidison"
date: "11 Apr 2016"
output: pdf_document
---

# Report

## Introduction

Looking at a data set of a collection of cars, we are interested in exploring the relationship between a set of variables and miles per gallon (MPG). We are particularly interested in the following two questions:

- “Is an automatic or manual transmission better for MPG?”
- "Quantifying the MPG difference between automatic and manual transmissions"


## Exploratory data analysis

First load the needed packages which are datasets, ggplot2 and gridExtra.

```{r, echo = FALSE, message = FALSE, results = 'hide'}
packages <- c("datasets", "ggplot2", "gridExtra")
sapply(packages, require, character.only = TRUE, quietly = TRUE)
```

Load the dataset from the `datasets` package and look at the basic information.

```{r, echo = FALSE, message = FALSE}
data(mtcars); head(mtcars, 3)
```

Look at the visual relationships of weight and acceleration on mileage per gallon according to transmission on Fig. 0 *(cf. Appendix Fig. 0)*.

```{r fig0, echo = FALSE, fig.keep = 'none'}
wtPlot <- ggplot(mtcars, aes(x = wt, y = mpg)) +
    geom_point(aes(color = am)) +
    labs(title = "Miles/gallon versus weight \n according to transmission",
         x = "Weight (in K lbs)",
         y = "Miles/gallon") +
    scale_fill_discrete(guide = FALSE) +
    theme(text = element_text(size = 7))

qsecPlot <- ggplot(mtcars, aes(x = qsec, y = mpg)) +
    geom_point(aes(color = am)) +
    labs(title = "Miles/gallon versus 1/4 mile time \n according to transmission",
         x = "Time to reach 1/4 mile (in sec)",
         y = "Miles/gallon") +
    scale_fill_discrete(guide = FALSE) +
    theme(text = element_text(size = 7))

grid.arrange(wtPlot, qsecPlot, ncol = 2)
```

We convert `am` into a factor; it is the variable we are interested in.

```{r}
mtcars$am <- factor(mtcars$am, labels = c("Auto", "Manual"))
```

And we plot transmission versus miles per gallon *(cf. Appendix Fig. 1)* to highlight any suspect behavior.

```{r fig1, echo = FALSE, fig.keep = 'none'}
ggplot(mtcars, aes(x = am, y = mpg, group = am)) +
    geom_boxplot() +
    labs(title = "Fig. 1: Miles per gallon versus type of transmission",
         x = "Transmission",
         y = "Miles/gallon") +
    scale_fill_discrete(guide = FALSE) +
    theme(text = element_text(size = 7))
```


## Statistical inference

From Fig. 1, the manual transmission seems to allow more mileage per gallon than the automatic transmission. We use a t-test.

```{r}
tTest <- t.test(mpg ~ am, data = mtcars)
```

```{r, echo = FALSE, message = FALSE}
print(paste0("P-value: ", tTest$p.value))
print(paste0(tTest$conf.int))
```

The p value is statistically significant (< 0.05). Plus, 0 is not in the confidence interval. We reject the Null hypothesis and confirm that automatic transmission allows less mileage per gallon than manual one **without considering any other variable**.


## Regression model

We use the linear regression to fit a model, starting with a backward removal method for building the regression model.

```{r}
stepModel <- summary(step(lm(mpg ~ ., mtcars), trace = 0))
```

```{r, echo = FALSE, message = FALSE}
print(paste0("Model: ", stepModel$call[2]))
print(paste0("Adjusted R-squared: ", stepModel$adj.r.squared))
```

This method tells us to keep the weight, the quarter mile time and thankfully the transmission. The model is not bad with an adjusted R-squared of 83%.

Test the model to optimize it by adding interaction from the transmission.

```{r}
fineModel <- lm(mpg ~ (am*wt) + qsec, mtcars)
```

```{r, echo = FALSE, message = FALSE}
summary(fineModel)$coef
print(paste0("Adjusted R-squared: ", summary(fineModel)$adj.r.squared))
```

We reached a 88% adjusted R-squared that we will check to avoid bias thanks to the residual plots *(cf. Appendix Fig. 2)*.

```{r fig2, echo = FALSE, fig.keep = 'none'}
par(mfrow = c(2,2))
plot(fineModel)
```

What we can say about the plots is that: the Residual vs. Fitted plot shows homoscedasticity that justifies the independance assumption; the residuals are approximately normally distributed according to the Normal Q-Q plot; the line representing the variance is rather constant on the Scale-Location plot; and everything is fine in the Residuals vs. Leverage plot.


## Results

### Conclusions

To answer the two questions initally asked:

**Is an automatic or manual transmission better for MPG?**
We can conclude that choosing the best transmission for mileage per gallon depends on the weight and acceleration of a car.

**Quantify the MPG difference between automatic and manual transmissions**
Choosing a manual transmission increases the mpg (+10.64 to +17.51) and so does the automatic transmission (+3.82 to +15.62). But the heavier the car, the more mpg decreases - -3.61 to -2.27 every 1,000 lbs - and even more with a manual transmission with -5.34 to -2.94 every 1,000 lbs. And the longer the quarter mile time, the slightly more mpg - +0.76 to +1.27 per second whatever the transmission.

That means that the heavier the cars with a quick acceleration the better thinking about an automatic transmission for better fuel efficiency. Small light cars would have a little bit more efficiency with manual transmission.

\pagebreak
  
# Appendix

Find figures related to the Report part.

## Fig. 0

```{r, ref.label = 'fig0', results = 'markup', fig.align = 'center'}
```

## Fig. 1

```{r, ref.label = 'fig1', results = 'markup', fig.align = 'center', fig.width = 5}
```

## Fig. 2

```{r, ref.label = 'fig2', results = 'markup', fig.align = 'center', fig.height = 6}
```

